HMM

### 2015-12-28 paper3.Rnw
more tweaks to methods

\cite{kerk2015hidden} found that the best-fitting HSMMs
according to BIC incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}.As \citeauthor{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of behavioural modes.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission (FWC) staff using
trained hounds and houndsmen. Locations used in this study were the
hourly GPS coordinates then decomposed into step lengths (in meters)
and turning angles (in radians).

Hidden semi-Markov models (HSMM), an extension of HMM that permits
explicit modelling of dwell times (ie. Time spent in a specific state)
using alternative distributions \citep{langrock_flexible_2012} were
used to reveal multiphasic movement of Florida panthers
\citep{kerk2015hidden}. Of 18 Florida panthers GPS-tracked in
hourly intervals, 13 panthers were modelled independently with 2 to 6
states time homogeneous HSMM model with alternative distributions for
step lengths (log-Normal, gamma, and Weibull), turning angle (von
Mises and wrapped Cauchy) and dwell times (Poisson and negative
binomial). Weibull and wrapped Cauchy were the choice of distributions
chosen via Akaike information criterion (AIC) among the alternative
distributions for step length and turning angle
respectively. Furthermore, Bayesian information criterion (BIC) were
used to evaluate goodness of fit between models with different number
of states. BIC values of these models decrease as the number of states
increased from three to six states suggesting the six states model was
favoured statistically, on the other hand, it was hard to interpret
the six states biologically. Despite the flexibility to model dwell
time, it did not affect results, thus, in this paper, we will use HMM
instead of HSMM. Furthermore, the state occupancies based on the
Viterbi algorithm, which predicts the most likely sequence of movement
states based on observed movement results
\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
future movement patterns and cannot capture which movement modes
panthers are likely to occupy depending on time of day from the time
homogeneous transitions (ie, the transition between all states are
constant through time).

@@

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting. BIC was used for model selection because we were interested in explaining the data generation process by identifying the 'true' number of underlying movement states.

We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by Morales et al. (2004). For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:

$\pi_{ij} = \frac{\mathrm{exp}(\eta_{ij}(t))}{1+\sum_{j=2}^{n}\mathrm{exp}(\eta_{ij}(t))},$ for $j={2,...,n}$

@@

$\eta_{ij}(t)= b_{ij1}+b_{ij2}\mathrm{cos}(\frac{2\pi t}{24})+b_{ij3}\mathrm{sin}(\frac{2\pi t}{24})$

@@

We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches. First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\bigtriangleup$B = BIC - min(BIC)). Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity. Lastly, autocorrelation comparison between the observed and simulated data.

@@

As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.

>>>>

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}.As \citeauthor{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of behavioural modes.

@@

\subsection{Data and previous analyses}

\todo{what do we think of this subsection title?}

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission (FWC) staff using
trained hounds and houndsmen. Of these panthers, 13 had 
sufficient data to be used by \cite{kerk2015hidden}. Here we focus
on the four cats with the most data, in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is to larger data sets \todo{give cat IDs - these are shown in Fig S2a of vdK.  Give threshold amount of data; justify/describe how we decided.  Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?}
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions.  As shown 
\todo{I was going to say ``but not emphasized'', but that seems snarky}
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work}

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.

Our analysis again aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model; we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the relatively subtle distinctions in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations. (Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations and neglecting possible correlations
between them.)  However, we do briefly compare all three
step-length distributions, with and without incorporating
a von Mises-distributed turning angle in the model (Figure~1).

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
suggesting that the six-state model was
favoured statistically. On the other hand, it was hard to interpret
the six states biologically.
We follow their lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the 'true' number 
of underlying movement states.  Using BIC also simplifies \ldots
\todo{cite S. Richards Ecology paper?  it's easier to see if BIC
identifies the true model than to see if AIC finds the model with
the minimum K-L distance; admit that tapering effect sizes do
happen here; claim similar results with AIC, although
generally higher number of states identified}

\todo{This description of the V. algorithm is not coherent; need to give
just a \emph{brief} definition here (since we are going to use it) and criticize/
explain shortcomings later}

Furthermore, the state occupancies based on the
Viterbi algorithm, which predicts the most likely sequence of movement
states based on observed movement results
\citep{zucchini_hidden_2009,langrock_flexible_2012} cannot predict
future movement patterns and cannot capture which movement modes
panthers are likely to occupy depending on time of day from the time
homogeneous transitions (ie, the transition between all states are
constant through time).

\subsection{Model description}

@@

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.

We modeled three types of transition complexity instead of time homogeneous transition; extension of "Switch with covariates" by \cite{morales_extracting_2004}. For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:

$\pi_{ij} = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right),$ for $j={2,...,n}$

@@

$\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})$

@@

We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches. First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\Delta$ BIC = BIC - min(BIC)). Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity. Lastly, autocorrelation comparison between the observed and simulated data.

@@

As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.
